---
title: "Karyological data analysis"
author: "Manuel Tiburtini"
date: '2023-01-20'
output: html_document
---

# Introduction

This code is part of the karyological data analysis conducted in Tiburtini et al., 2023 link.

In this section, karyological asymetry data measured on 91 seeds individual from 14 populations are being analyzed.

# Data explanation

The raw data set consists in 91 individuals x 4 karyological indices. Moreover, position of the centromere and presence of satellites are listed. Factor includes COD_LOC mean the locality acronym name, SPECIE to species level and TAXON include subspecies rank.

### Features description

1.  THL : Total Haploid Length.
2.  CVCI: measuring how heterogeneous are centromeres positions in a karyotype.
3.  CVCL: quantifying interchromosomal and karyotype asymmetry,
4.  MCA: quantifying intrachromosomal karyotype asymmetry.
5.  m : number of metacentric chromosomes.
6.  sm : number of submetacentric chromosomes.
7.  st : number of subtelocentric chromosomes.
8.  satelliti : presence of satellites on the short arm.

```{r setup, data import and cleaning}
library(tidyverse)

karyo <-  read.csv("/Users/manueltiburtini/Desktop/Dottorato/Armeria\ sardo-Corsa/Analisi\ dati\ Tiburtini\ et\ al.\ 2023/Karyology/karyo.csv", sep=",", stringsAsFactors = TRUE)

karyo <- karyo %>% select(-COD_CAGL,-note,-X,-X.1) 

karyo
```

```{r descriptive statistics}

#Descriptive statistics by COD_LOC: mean and standard deviation
karyo %>% 
    select(-TAXON,-SPECIE,-HYP_5, -Vetrino,-Piastra,-x,-X2n, COD_LOC,THL,CVCI,CVCL,MCA,-m,-sm,-st,-satelliti) %>% 
  group_by(COD_LOC) %>% 
  reframe(across(everything(), 
                 list(comb = ~ paste(sprintf("%.2f", mean(.x)), 
                                     sprintf("%.2f", sd(.x)), sep = " ± ")))) %>%
  pivot_longer(cols = -COD_LOC, names_to = c(".value", "variable"), 
               names_sep = "_") %>% 
  rename_with(~ str_remove(., "_comb"), contains("_comb")) %>% 
  as.tibble() %>% select(-variable)

#Descriptive statistics by COD_LOC:  sample size
karyo %>% 
  select(-TAXON,-SPECIE,COD_LOC,THL,CVCI,CVCL,MCA) %>% 
  group_by(COD_LOC) %>% 
  summarize(n=n()) %>% 
  write.csv(., "descriptive_karyo_stat.csv")

#frequency of chromosome type
karyo %>% 
  select(m,sm,st) %>% 
  summarize(sum_m=sum(m),sum_sm=sum(sm),sum_st=sum(st), n=sum(sum_m,sum_sm,sum_st)) %>% 
  mutate(freq_m = sum_m / sum(n),freq_sm = sum_sm / sum(n),freq_st = sum_st / sum(n))

#number of satellites
karyo %>% 
  select(COD_LOC, satelliti) %>% 
  filter(satelliti=="SI")

#mean size
karyo %>% 
  select(THL,x) %>% 
  mutate(size=THL/x) %>% 
  summarize(mean=mean(size), dev.st=sd(size))

```

```{r normality test}

karyo %>%
  select(-TAXON,-SPECIE,-HYP_5, -Vetrino,-Piastra,-x,-X2n, COD_LOC,THL,CVCI,CVCL,MCA,-m,-sm,-st,-satelliti) %>% 
  group_by(COD_LOC) %>% 
  summarise(sw_THL = rstatix::shapiro_test(THL)$p.value,
            sw_CVCI = rstatix::shapiro_test(CVCI)$p.value,
            sw_CVCL = rstatix::shapiro_test(CVCL)$p.value,
            sw_MCA = rstatix::shapiro_test(MCA)$p.value) %>% 
  filter(sw_THL <0.05 |sw_CVCI <0.05 | sw_CVCL <0.05 |sw_MCA <0.05) 

```

```{r MCA-CVCL scatterplot}
library(ggforce)
library(ggplot2)

ggplot(karyo, aes(MCA, CVCL, color=SPECIE,shape=COD_LOC)) + geom_point(size=2)+
geom_mark_hull(aes(color = SPECIE , group=SPECIE),expand = unit(0, "mm"),radius = unit(0, "mm"), concavity=20)+
   scale_shape_manual(values = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15)) +
  scale_color_manual(values = c("#E84D6B", "#EEBA0B", "#FF8E3C", "#C59ED8", "#AFD958", "#5BAFD9")) +
  theme_bw() +
ylab(bquote(CV[CL]))+xlab(bquote(M[CA]))+ theme(axis.text.x = element_text(size = 10),axis.text.y = element_text(size = 10)) # + stat_ellipse(type = "t", linetype = 2, level = 0.99)
```

```{r univariate pairwise permutation test }
library(rcompanion)

alpha <- 0.01
#for subspecies
karyo %>% 
  reframe(Comparison=pairwisePermutationTest(CVCL ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$Comparison,
    THL=pairwisePermutationTest(THL ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust, 
            CVCL=pairwisePermutationTest(CVCL ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            CVCI=pairwisePermutationTest(MCA ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            MCA=pairwisePermutationTest(MCA ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust) %>% 
filter(THL < alpha |CVCI < alpha | CVCL < alpha |MCA < alpha)
  
#per taxon
karyo_subset %>% 
  reframe(Comparison=pairwisePermutationTest(CVCL ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$Comparison,
    THL=pairwisePermutationTest(THL ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust, 
            CVCL=pairwisePermutationTest(CVCL ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            CVCI=pairwisePermutationTest(MCA ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            MCA=pairwisePermutationTest(MCA ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust) %>% 
filter(THL <alpha |CVCI <alpha | CVCL <alpha |MCA <alpha)

#ALPHA 0.05
#for subspecies

karyo %>% 
  reframe(Comparison=pairwisePermutationTest(CVCL ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$Comparison,
    THL=pairwisePermutationTest(THL ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust, 
            CVCL=pairwisePermutationTest(CVCL ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            CVCI=pairwisePermutationTest(MCA ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            MCA=pairwisePermutationTest(MCA ~ TAXON, data   = ., method = "BH",  alternative="two.sided")$p.adjust) %>% filter(THL <0.05 |CVCI <0.05 | CVCL <0.05 |MCA <0.05) %>% 
    write.csv(., "significance_katyo.csv")
  
#per species
karyo_subset %>% 
  reframe(Comparison=pairwisePermutationTest(CVCL ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$Comparison,
    THL=pairwisePermutationTest(THL ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust, 
            CVCL=pairwisePermutationTest(CVCL ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            CVCI=pairwisePermutationTest(MCA ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            MCA=pairwisePermutationTest(MCA ~ SPECIE, data   = ., method = "BH",  alternative="two.sided")$p.adjust) %>% 
filter(THL <0.05 |CVCI <0.05 | CVCL <0.05 |MCA <0.05)

#new circumscriptions
karyo %>% 
  reframe(Comparison=pairwisePermutationTest(CVCL ~ HYP_5, data   = ., method = "BH",  alternative="two.sided")$Comparison,
    THL=pairwisePermutationTest(THL ~ HYP_5, data   = ., method = "BH",  alternative="two.sided")$p.adjust, 
            CVCL=pairwisePermutationTest(CVCL ~ HYP_5, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            CVCI=pairwisePermutationTest(MCA ~ HYP_5, data   = ., method = "BH",  alternative="two.sided")$p.adjust,
            MCA=pairwisePermutationTest(MCA ~ HYP_5, data   = ., method = "BH",  alternative="two.sided")$p.adjust) %>% filter(THL <alpha |CVCI <alpha | CVCL <alpha |MCA <alpha) 

```
